Predator clusters

### 2023-06-27 reports/01_predators.qmd
Update species maps

## Antarctic Peninsula winter predators

### Counts

The five most abundant species (in ind nmi^-1).

@@

### Maps

@@

@@

effort_stars <- effort_sf %>% 
  filter(nmi < 1000) %>% 
  rasterize_counts(count_col = "nmi",
                   res = 2.5e4)
map_lim <- st_bbox(c(xmin = -65, xmax = -55, ymin = -65, ymax = -58),
                       crs = "EPSG:4326") %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)
ant_basemap(map_lim) + 
  stars::geom_stars(aes(fill = nmi), data = effort_stars) +
  scale_fill_distiller(palette = "PuRd", 
                       direction = 1, 
                       na.value = NA) +
  scale_x_continuous(breaks = seq(-65, -55, by = 5)) +
  scale_y_continuous(breaks = seq(-65, -58, by = 2)) +
  coord_ant(map_lim) +
  facet_wrap(vars(year)) +
  expand_limits(fill = 0)

@@

Distribution maps of the five most abundant species in each survey year, aggregated to 25km bins.

@@

topfive_sf <- predators_agg %>% 
  semi_join(top_five, by = "species") %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))

topfive_stars <- rasterize_counts(topfive_sf, res = 2.5e4)

map_lim <- st_bbox(c(xmin = -65, xmax = -55, ymin = -66, ymax = -60),
                       crs = "EPSG:4326") %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.1)

ant_basemap(map_lim) +
  stars::geom_stars(aes(fill = log(count_norm + 1)), 
                    data = topfive_stars) +
  scale_fill_distiller(palette = "PuRd",
                       direction = 1,
                       na.value = NA) +
  scale_x_continuous(breaks = c(-65, -60, -55)) +
  scale_y_continuous(breaks = c(-66, -64, -62, -60)) +
  coord_ant(map_lim) +
  facet_grid(species ~ year) +
  expand_limits(fill = 0) +
  theme(axis.text.x = element_text(size = 7))
```

@@

soi_stars <- rasterize_counts(soi_sf, res = 2.5e4)

map_lim <- st_bbox(soi_sf) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

ant_basemap(map_lim) +
  stars::geom_stars(aes(fill = log(count_norm + 1)), data = soi_stars) +
  scale_fill_distiller(palette = "PuRd",
                       direction = 1,
                       na.value = NA) +
  scale_x_continuous(breaks = c(-65, -60, -55)) +
  scale_y_continuous(breaks = c(-64, -62, -60, -58)) +
  facet_grid(species ~ year) +
  expand_limits(fill = 0) +
  coord_ant(map_lim) +
  theme(axis.text.x = element_text(size = 7))
```

>>>>

## Counts

The five most abundant species (in ind nmi\^-1).

@@

## Maps

@@

### Climatologies

```{r}
#| label: species_climatologies

library(terra)
# Mkae "top five" predator sightings spatially aware
topfive_sf <- predators_agg %>% 
  semi_join(top_five, by = "species") %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))
# Create a template raster of the study area
effort_ext <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  ext()
rast_template <- rast(crs = as.character(ant_proj())[[1]],
                      extent = effort_ext,
                      resolution = 30e3)
# Rasterize effort by year
effort_rast <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  rasterize(rast_template,
            field = "nmi",
            by = "year",
            fun = sum)
# Rasterize the predator sightings by year and species
topfive_rast <- topfive_sf %>% 
  st_transform(ant_proj()) %>% 
  mutate(year_species = interaction(year, species)) %>% 
  rasterize(rast_template, 
            field = "count",
            by = "year_species",
            fun = sum)
# Normalize counts by effort
topfive_norm <- topfive_rast
for (i in seq(nlyr(topfive_rast))) {
  lyr_name <- names(topfive_rast)[i]
  lyr_year <- substr(lyr_name, 1, 4)
  effort_year <- effort_rast[lyr_year] %>% 
    subst(0, NA)
  topfive_norm[[i]] <- topfive_norm[[i]] / effort_year
}
# Average across years
topfive_climatology <- unique(top_five$species) %>% 
  map(function(sp) {
    sp_rast <- subset(topfive_norm, 
                      which(str_detect(names(topfive_norm), sp)))
    mean(sp_rast, na.rm = TRUE)
  }) %>% 
  set_names(unique(top_five$species)) %>% 
  rast()
# Classify raster (otherwise CRSE throw it all off)
topfive_clim_class <- topfive_climatology %>% 
  classify(rcl = c(0, 1, 5, 10, 50))

# Project bounding box
map_bbox <- ext(-63, -54, -64, -59) %>% 
  project(from = "+proj=longlat +datum=WGS84",
          to = as.character(ant_proj())[[1]])

# Make the map! 
ant_basemap() +
  tidyterra::geom_spatraster(data = topfive_clim_class) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_wrap(~lyr) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "top")

### Survey effort

@@

map_bbox <- ext(effort_rast)

ant_basemap(map_bbox) +
  tidyterra::geom_spatraster(data = effort_rast) +
  scale_fill_distiller(palette = "PuRd", 
                       direction = 1, 
                       na.value = NA) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -57, by = 2)) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  labs(fill = "nmi") +
  facet_wrap(~lyr) +
  expand_limits(fill = 0)

@@

### Yearly distributions

Distribution maps of the five most abundant species in each survey year, aggregated to 30km bins.

@@

map_bbox <- ext(-63, -54, -64, -59) %>% 
  project(from = "+proj=longlat +datum=WGS84",
          to = as.character(ant_proj())[[1]])

topfive_class <- topfive_rast %>% 
  classify(rcl = c(0, 1, 5, 10, 50, 100, 1000, Inf))

topfive_class_df <- as.data.frame(topfive_class, xy = TRUE) %>% 
  pivot_longer(-c("x", "y"), 
               names_to = "name", 
               values_to = "count") %>% 
  separate_wider_delim(name, 
                       delim = ".", 
                       names = c("year", "species")) %>% 
  drop_na(count)

ant_basemap() +
  geom_raster(aes(x, y, fill = count), topfive_class_df) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_grid(species ~ year) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "right",
        text = element_text(size = 8))
```

@@

soi_rast <- soi_sf %>% 
  st_transform(ant_proj()) %>% 
  mutate(year_species = interaction(year, species)) %>% 
  rasterize(rast_template, 
            field = "count",
            by = "year_species",
            fun = sum)

map_bbox <- ext(-63, -54, -64, -59) %>% 
  project(from = "+proj=longlat +datum=WGS84",
          to = as.character(ant_proj())[[1]])

soi_class <- soi_rast %>% 
  classify(rcl = c(0, 1, 5, 10, 50, 100, 1000))

soi_class_df <- as.data.frame(soi_class, xy = TRUE) %>% 
  pivot_longer(-c("x", "y"), 
               names_to = "name", 
               values_to = "count") %>% 
  separate_wider_delim(name, 
                       delim = ".", 
                       names = c("year", "species")) %>% 
  drop_na(count)

ant_basemap() +
  geom_raster(aes(x, y, fill = count), soi_class_df) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_grid(species ~ year) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "right",
        text = element_text(size = 8))
```

### 2023-06-27 reports/04_pred_clus.qmd
Update species maps

>>>>

### Species within clusters

```{r}
#| label: clusters_species
predators_abundant %>% left_join(
  tibble(
    amlr.station = names(clust),
    pred_clust = factor(clust,
                        labels = c("Open", "Marginal", "Pagophilic"))
  ), by = "amlr.station") %>% 
  complete(amlr.station, pred_clust, species, 
           fill = list(count_norm = 0)) %>% 
  group_by(species, pred_clust) %>% 
  summarize(mean_abundance = mean(count_norm), .groups = "drop") %>%
  mutate(species = fct_reorder(species, 
                               mean_abundance, 
                               .fun = sum,
                               .desc = TRUE)) %>% 
  ggplot(aes(species, mean_abundance)) +
  geom_col() +
  labs(y = "Mean abundance (ind nmi^-1)") +
  facet_wrap(~ pred_clust, ncol = 1) +
  theme_classic(base_size = 14) +
  theme(axis.title.x = element_blank())
```
